Orienting the causal relationship between imprecisely measured traits using GWAS summary data

### 2017-11-13 docs/manuscript-plos.tex
updated

\setlength{\droptitle}{-2em}
  \title{Orienting the causal relationship between imprecisely measured traits
using GWAS summary data}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{24 October 2017}

@@

\subsubsection{Abstract}\label{abstract}

@@

\subsubsection{Author summary}\label{author-summary}

@@

\subsection{Introduction}\label{introduction}

@@

\subsection{Methods}\label{methods}

\subsubsection{CIT test}\label{cit-test}

First we describe how the CIT method {[}4{]} is implemented in the R
package \emph{R/cit} {[}18{]}. The methodology of the CIT is as follows.
Assume an exposure \(x\) is instrumented by a SNP \(g\), and the
exposure \(x\) causes an outcome \(y\), as described above. The
following tests are then performed:

@@

The term in the 4th test can be rewritten as a
\(cov(g, y|x) = cov(g, y - \hat{y})\) where
\(y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x\) is the residual
of \(y\) after adjusting for \(x\), and \(x\) is assumed to mediate the
association between the SNP and the outcome. The condition in the 4th
test is formulated as an equivalence testing problem that is estimated
using simulations, comparing the estimate from the data against
empirically obtained estimates for simulated variables where the
independence model is true (full details are given in {[}4{]}). We note
here that this approach is liable to fail, even when there is a true
causal relationship, when confounders of the exposure and outcome are
present, as these will induce collider bias.

@@

For the purposes of compiling simulation results we use an arbitrary
\(\alpha = 0.05\) value, though we stress that for real analyses it is
not good practice to rely on p-values for making causal inference, nor
is it reliable to depend on arbitrary significance thresholds {[}39{]}.

\subsubsection{MR causal test}\label{mr-causal-test}

Two stage least squares (2SLS) is a commonly used technique for
performing MR when the exposure, outcome and instrument data are all
available in the same sample. A p-value for this test, \(p_{MR}\), was
obtained using the \(systemfit\) function in the R package
\(R/systemfit\) {[}40{]}. Note that the value of \(p_{MR}\) is identical
when using the same genetic variant to instrument the influence of the
exposure \(x\) on the outcome \(y\), or erroneously, instrumenting the
outcome \(y\) on the exposure \(x\).

@@

To infer the direction of causality it is desirable to know which of the
variables, \(x\) or \(y\), is being directly influenced by the
instrument \(g\). This can be achieved by assessing which of the two
variables has the biggest absolute correlation with \(g\) (S2 Text),
formalised by testing for a difference in the correlations \(\rho_{gx}\)
and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
within a population {[}41{]}. It is calculated as

@@

An advantage of using the Steiger test in the two sample context is that
it can compare correlations in independent samples where sample sizes
are different. Steiger test statistics were calculated using the
\emph{r.test} function in the R package \emph{R/psych} {[}42{]}.

@@

\subsubsection{Causal direction sensitivity analysis for measurement
error}\label{causal-direction-sensitivity-analysis-for-measurement-error}

@@

\subsubsection{Simulations}\label{simulations}

@@

Two sample MR {[}30{]} was performed using summary statistics for
genetic influences on gene expression and DNA methylation. To do this we
obtained a list of 458 gene expression - DNA methylation associations as
reported in Shakhbazov et al {[}43{]}. These were filtered to be located
on the same chromosome, have robust correlations after correcting for
multiple testing, and to share a SNP that had a robust cis-acting effect
on both the DNA methylation probe and the gene expression probe. Because
only summary statistics were available (effect, standard error, effect
allele, sample size, p-values) for the instrumental SNP on the
methylation and gene expression levels, the Steiger test of two
independent correlations was used to infer the direction of causality
for each of the associations. The Wald ratio test was then used to
estimate the causal effect size for the estimated direction for each
association.

All analysis was performed using the R programming language {[}44{]} and
code is made available at
\url{https://github.com/explodecomputer/causal-directions} and
implemented in the MR-Base (\url{http://wwww.mrbase.org}) platform
{[}26{]}.

\subsection{Results}\label{results}

\subsubsection{Mediation-based causal inference under measurement
error}\label{mediation-based-causal-inference-under-measurement-error}

\subsubsection{Using MR Steiger to infer the direction of
causality}\label{using-mr-steiger-to-infer-the-direction-of-causality}

\subsubsection{Comparison of CIT and MR Steiger for obtaining the
correct direction of
causality}\label{comparison-of-cit-and-mr-steiger-for-obtaining-the-correct-direction-of-causality}

\subsubsection{The causal relationship between gene expression and DNA
methylation
levels}\label{the-causal-relationship-between-gene-expression-and-dna-methylation-levels}

We performed two sample MR {[}30{]} for each association in the
direction of causality inferred by the Stieger test. We observed that
the sign of the MR estimate was generally in the same direction as the
Pearson correlation coefficient reported by Shakhbazov et al {[}43{]}
(Figure 6b). There was a moderate correlation between the absolute
magnitudes of the causal correlation and the observational Pearson
correlation (r = 0.45). Together these inferences suggest that even in
estimating associations between `omic' variables, which are considered
to be low level phenotypes, it is important to use causal inference
methods over observational associations to infer causal effect sizes.

We also observed that for associations where methylation caused gene
expression the causal effect was more likely to be negative than for the
associations where gene expression caused methylation (OR = 0.61 (95\%
CI 0.36 - 1.03), Figure 6c), suggesting that reducing methylation levels
at a controlling CpG typically leads to increased gene expression
levels, consistent with expectation {[}45{]}.

\subsection{Discussion}\label{discussion}

Researchers are often confronted with the problem of making causal
inferences using a statistical framework on observational data. In the
epidemiological literature issues of measurement error in mediation
analysis are relatively well explored {[}46{]}. Our analysis extends
this to related methods such as CIT that are used in predominantly 'omic
data. These methods are indeed susceptible to the same problem as
standard mediation based analysis, and specifically we show that as
measurement error in the (true) exposure variable increases, CIT is
likely to have reduced statistical power, and liable to infer the wrong
direction of causality. We also demonstrate that, though unintuitive,
increasing sample size does not resolve the issue, rather it leads to
more extreme p-values for the model that predicts the wrong direction of
causality.

We demonstrate this new method by evaluating the causal relationships of
458 known associations between DNA methylation and gene expression
levels using summary level data. The inferred causal direction is
heavily influenced by how much measurement error is present in the
different assaying platforms. For example, if DNA methylation measures
typically have higher measuremet error than gene expression measures
then our analysis suggests that DNA methylation levels would be more
often the causal factor in the association. Indeed, previous studies
which have evaluated measurement error in these platforms do support
this position {[}47,48{]}, though making strong conclusions for this
analysis is difficult because measurement error is likely to be study
specific. We also haven't accounted for the influence of winner's curse,
which can inflate estimates of the variance explained by SNPs, with
higher inflation expected amongst lower powered studies. Using p-values
for genetic associations from replication studies will mitigate this
problem.

In our simulations we focused on the simple case of a single instrument
in a single sample setting with a view to making a fair comparison
between MR and the various mediation-based methods available. However,
if there is only a single instrument it is difficult to separate between
the two competing models of \(g\) instrumenting a trait which causes
another trait, and \(g\) having pleiotropic effects on both traits
independently {[}49{]}. Under certain conditions of measurement error
the CIT test can distinguish these models. We also note that it is
straightforward to extend the MR Steiger approach to multiple
instruments, requiring only that the total variance explained by all
instruments be calculated under the assumption that they are
independent. Multiple instruments can indeed help to distinguish between
the causal and pleiotropic models, for example by evaluating the
proportionality of the SNP-exposure and SNP-outcome effects {[}16{]}.
Additionally, if there is at least one instrument for each trait then
bi-directional MR can offer solutions to inferring the causal direction
{[}16,28,50{]}. We restricted the simulations to evaluating the causal
inference between quantitative traits, but it is possible that the
analysis could be extended to binary traits by using the genetic
variance explained on the liability scale, taking into account the
population prevalence {[}51{]}. However, our analysis goes beyond many
previous explorations of measurement error by assessing the impacts of
both imprecision (noise) and linear transformations of the true variable
on causal inference.

Our new method attempts to infer causal directions under the assumption
that horizontal pleiotropy (the influence of the instrument on the
outcome through a mechanism other than the exposure) is not present.
Recent method developments in MR {[}24,25{]} have focused on accounting
for the issues that horizontal pleiotropy can introduce when multiple
instruments are available, but how they perform in the presence of
measurement error remains to be explored. An important advantage that MR
confers over most mediation based analysis is that it can be performed
in two samples, which can considerably improve power and expand the
scope of analysis. However, whether there is a substantive difference in
two sample MR versus one sample MR in how measurement error has an
effect is not yet fully understood. We have also assumed no measurement
error in the genetic instrument, which is not unreasonable given the
strict QC protocols that ensure high quality genotype data is available
to most studies. We have restricted the scope to only exploring
non-differential measurement error and avoided the complications
incurred if measurement error in the exposure and outcome is correlated.
We have also not addressed other issues pertaining to instrumental
variables which are relevant to the question of instrument-exposure
specification. One such problem is exposure misspecification, for
example an instrument could associate with several closely related
putative outcomes, with only one of them actually having a causal effect
on the outcome. This problem has shown to be the case for SNPs
influencing different lipid fractions, for example {[}52,53{]}.

Mediation based network approaches, that go beyond analyses of two
variables, are very well established {[}37{]} and have a number of
extensions that make them valuable tools, including for example network
construction. But because they are predicated on the basic underlying
principles of mediation they are liable to suffer from the same issues
of measurement error. Recent advances in MR methodology, for example
applying MR to genetical genomics {[}54{]}, multivariate MR {[}53{]} and
mediation through MR {[}55--57{]} may offer more robust alternatives for
these more complicated problems.

\subsection{Figures}\label{figures}

@@

\hypertarget{ref-Sterne2001}{}
39. Sterne JAC, Smith GD. Sifting the evidence---what's wrong with
significance tests? BMJ. 2001;322: 226--231.

\hypertarget{ref-Henningsen2007}{}
40. Henningsen A, Hamann JD. systemfit : A Package for Estimating
Systems of Simultaneous Equations in R. Journal of Statistical Software.
2007;23: 1--40.
doi:\href{https://doi.org/10.18637/jss.v023.i04}{10.18637/jss.v023.i04}

\hypertarget{ref-Steiger1980}{}
41. Steiger JH. Tests for comparing elements of a correlation matrix.
Psychological Bulletin. 1980;87: 245--251.
doi:\href{https://doi.org/10.1037/0033-2909.87.2.245}{10.1037/0033-2909.87.2.245}

\hypertarget{ref-Revelle2015}{}
42. Revelle W. psych: Procedures for Psychological, Psychometric, and
Personality Research {[}Internet{]}. Evanston, Illinois: Northwestern
University; 2015. Available:
\url{http://cran.r-project.org/package=psych}

\hypertarget{ref-Shakhbazov2016}{}
43. Shakhbazov K, Powell JE, Hemani G, Henders AK, Martin NG, Visscher
PM, et al. Shared genetic control of expression and methylation in
peripheral blood. BMC genomics. BioMed Central; 2016;17: 278.
doi:\href{https://doi.org/10.1186/s12864-016-2498-4}{10.1186/s12864-016-2498-4}

\hypertarget{ref-RCoreTeam2015}{}
44. R Core Team. R: A Language and Environment for Statistical Computing
{[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
2015. Available: \url{https://www.r-project.org/}

\hypertarget{ref-Bird2002}{}
45. Bird A. DNA methylation patterns and epigenetic memory. Genes \&
development. 2002;16: 6--21.
doi:\href{https://doi.org/10.1101/gad.947102}{10.1101/gad.947102}

\hypertarget{ref-Cole2014}{}
46. Cole DA, Preacher KJ. Manifest Variable Path Analysis: Potentially
Serious and Misleading Consequences Due to Uncorrected Measurement
Error. Psychological Methods. 2014;19: 300--315.
doi:\href{https://doi.org/10.1037/a0033805}{10.1037/a0033805}

\hypertarget{ref-Bose2014}{}
47. Bose M, Wu C, Pankow JS, Demerath EW, Bressler J, Fornage M, et al.
Evaluation of microarray-based DNA methylation measurement using
technical replicates: the Atherosclerosis Risk In Communities (ARIC)
Study. BMC Bioinformatics. 2014;15: 312.
doi:\href{https://doi.org/10.1186/1471-2105-15-312}{10.1186/1471-2105-15-312}

\hypertarget{ref-Bryant2011}{}
48. Bryant PA, Smyth GK, Robins-Browne R, Curtis N, Novak J, Sladek R,
et al. Technical Variability Is Greater than Biological Variability in a
Microarray Experiment but Both Are Outweighed by Changes Induced by
Stimulation. Khodursky AB, editor. PLoS ONE. Public Library of Science;
2011;6: e19556.
doi:\href{https://doi.org/10.1371/journal.pone.0019556}{10.1371/journal.pone.0019556}

\hypertarget{ref-Zhu2016}{}
49. Zhu Z, Zhang F, Hu H, Bakshi A, Robinson MR, Powell JE, et al.
Integration of summary data from GWAS and eQTL studies predicts complex
trait gene targets. Nature Genetics. Nature Research; 2016;48: 481--487.
doi:\href{https://doi.org/10.1038/ng.3538}{10.1038/ng.3538}

\hypertarget{ref-Mancuso2017}{}
50. Mancuso N, Shi H, Goddard P, Kichaev G, Gusev A, Pasaniuc B.
Integrating Gene Expression with Summary Association Statistics to
Identify Genes Associated with 30 Complex Traits. The American Journal
of Human Genetics. 2017;100: 473--487.
doi:\href{https://doi.org/10.1016/j.ajhg.2017.01.031}{10.1016/j.ajhg.2017.01.031}

\hypertarget{ref-Lee2013c}{}
51. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
association studies: quantification of power and genomic prediction
accuracy. PLoS One. 2013;8: e71494.
doi:\href{https://doi.org/10.1371/journal.pone.0071494\%20\%5Bdoi\%5D/rPONE-D-13-18776\%20\%5Bpii\%5D}{10.1371/journal.pone.0071494 {[}doi{]}\textbackslash{}rPONE-D-13-18776 {[}pii{]}}

\hypertarget{ref-Do2013}{}
52. Do R, Willer CJ, Schmidt EM, Sengupta S, Gao C, Peloso GM, et al.
Common variants associated with plasma triglycerides and risk for
coronary artery disease. Nature Genetics. Nature Research; 2013;45:
1345--1352. doi:\href{https://doi.org/10.1038/ng.2795}{10.1038/ng.2795}

\hypertarget{ref-Burgess2014a}{}
53. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
multivariable Mendelian randomization to disentangle the causal effects
of lipid fractions. PloS one. Public Library of Science; 2014;9:
e108891.
doi:\href{https://doi.org/10.1371/journal.pone.0108891}{10.1371/journal.pone.0108891}

\hypertarget{ref-Relton2012}{}
54. Relton CL, Davey Smith G. Two-step epigenetic Mendelian
randomization: a strategy for establishing the causal role of epigenetic
processes in pathways to disease. International journal of epidemiology.
2012;41: 161--76.
doi:\href{https://doi.org/10.1093/ije/dyr233}{10.1093/ije/dyr233}

\hypertarget{ref-Varbo2015}{}
55. Varbo A, Benn M, Smith GD, Timpson NJ, Tybjaerg-Hansen A,
Nordestgaard BG. Remnant cholesterol, low-density lipoprotein
cholesterol, and blood pressure as mediators from obesity to ischemic
heart disease. Circulation research. 2015;116: 665--73.
doi:\href{https://doi.org/10.1161/CIRCRESAHA.116.304846}{10.1161/CIRCRESAHA.116.304846}

\hypertarget{ref-Burgess2015}{}
56. Burgess S, Daniel RM, Butterworth AS, Thompson SG. Network Mendelian
randomization: using genetic variants as instrumental variables to
investigate mediation in causal pathways. International journal of
epidemiology. 2015;44: 484--95.
doi:\href{https://doi.org/10.1093/ije/dyu176}{10.1093/ije/dyu176}

\hypertarget{ref-Richmond2016}{}
57. Richmond RC, Hemani G, Tilling K, Davey Smith G, Relton CL.
Challenges and novel approaches for investigating molecular mediation.
Human molecular genetics. Oxford University Press; 2016;25: R149--R156.
doi:\href{https://doi.org/10.1093/hmg/ddw197}{10.1093/hmg/ddw197}

>>>>

\setlength{\droptitle}{-2em}
  \title{Orienting the causal relationship between imprecisely measured traits
using GWAS summary data}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{13 November 2017}

@@

\section{Abstract}\label{abstract}

@@

\section{Author summary}\label{author-summary}

@@

\section{Introduction}\label{introduction}

@@

\section{Results}\label{results}

\subsection{Mediation-based causal inference under measurement
error}\label{mediation-based-causal-inference-under-measurement-error}

\subsection{Using MR Steiger to infer the direction of
causality}\label{using-mr-steiger-to-infer-the-direction-of-causality}

\subsection{Comparison of CIT and MR Steiger for obtaining the
correct direction of
causality}\label{comparison-of-cit-and-mr-steiger-for-obtaining-the-correct-direction-of-causality}

\subsection{The causal relationship between gene expression and DNA
methylation
levels}\label{the-causal-relationship-between-gene-expression-and-dna-methylation-levels}

We performed two sample MR {[}30{]} for each association in the
direction of causality inferred by the Stieger test. We observed that
the sign of the MR estimate was generally in the same direction as the
Pearson correlation coefficient reported by Shakhbazov et al {[}39{]}
(Figure 6b). There was a moderate correlation between the absolute
magnitudes of the causal correlation and the observational Pearson
correlation (r = 0.45). Together these inferences suggest that even in
estimating associations between `omic' variables, which are considered
to be low level phenotypes, it is important to use causal inference
methods over observational associations to infer causal effect sizes.

We also observed that for associations where methylation caused gene
expression the causal effect was more likely to be negative than for the
associations where gene expression caused methylation (OR = 0.61 (95\%
CI 0.36 - 1.03), Figure 6c), suggesting that reducing methylation levels
at a controlling CpG typically leads to increased gene expression
levels, consistent with expectation {[}40{]}.

\section{Discussion}\label{discussion}

Researchers are often confronted with the problem of making causal
inferences using a statistical framework on observational data. In the
epidemiological literature issues of measurement error in mediation
analysis are relatively well explored {[}41{]}. Our analysis extends
this to related methods such as CIT that are used in predominantly 'omic
data. These methods are indeed susceptible to the same problem as
standard mediation based analysis, and specifically we show that as
measurement error in the (true) exposure variable increases, CIT is
likely to have reduced statistical power, and liable to infer the wrong
direction of causality. We also demonstrate that, though unintuitive,
increasing sample size does not resolve the issue, rather it leads to
more extreme p-values for the model that predicts the wrong direction of
causality.

We demonstrate this new method by evaluating the causal relationships of
458 known associations between DNA methylation and gene expression
levels using summary level data. The inferred causal direction is
heavily influenced by how much measurement error is present in the
different assaying platforms. For example, if DNA methylation measures
typically have higher measuremet error than gene expression measures
then our analysis suggests that DNA methylation levels would be more
often the causal factor in the association. Indeed, previous studies
which have evaluated measurement error in these platforms do support
this position {[}42,43{]}, though making strong conclusions for this
analysis is difficult because measurement error is likely to be study
specific. We also haven't accounted for the influence of winner's curse,
which can inflate estimates of the variance explained by SNPs, with
higher inflation expected amongst lower powered studies. Using p-values
for genetic associations from replication studies will mitigate this
problem.

In our simulations we focused on the simple case of a single instrument
in a single sample setting with a view to making a fair comparison
between MR and the various mediation-based methods available. However,
if there is only a single instrument it is difficult to separate between
the two competing models of \(g\) instrumenting a trait which causes
another trait, and \(g\) having pleiotropic effects on both traits
independently {[}44{]}. Under certain conditions of measurement error
the CIT test can distinguish these models. We also note that it is
straightforward to extend the MR Steiger approach to multiple
instruments, requiring only that the total variance explained by all
instruments be calculated under the assumption that they are
independent. Multiple instruments can indeed help to distinguish between
the causal and pleiotropic models, for example by evaluating the
proportionality of the SNP-exposure and SNP-outcome effects {[}16{]}.
Additionally, if there is at least one instrument for each trait then
bi-directional MR can offer solutions to inferring the causal direction
{[}16,28,45{]}. We restricted the simulations to evaluating the causal
inference between quantitative traits, but it is possible that the
analysis could be extended to binary traits by using the genetic
variance explained on the liability scale, taking into account the
population prevalence {[}46{]}. However, our analysis goes beyond many
previous explorations of measurement error by assessing the impacts of
both imprecision (noise) and linear transformations of the true variable
on causal inference.

Our new method attempts to infer causal directions under the assumption
that horizontal pleiotropy (the influence of the instrument on the
outcome through a mechanism other than the exposure) is not present.
Recent method developments in MR {[}24,25{]} have focused on accounting
for the issues that horizontal pleiotropy can introduce when multiple
instruments are available, but how they perform in the presence of
measurement error remains to be explored. An important advantage that MR
confers over most mediation based analysis is that it can be performed
in two samples, which can considerably improve power and expand the
scope of analysis. However, whether there is a substantive difference in
two sample MR versus one sample MR in how measurement error has an
effect is not yet fully understood. We have also assumed no measurement
error in the genetic instrument, which is not unreasonable given the
strict QC protocols that ensure high quality genotype data is available
to most studies. We have restricted the scope to only exploring
non-differential measurement error and avoided the complications
incurred if measurement error in the exposure and outcome is correlated.
We have also not addressed other issues pertaining to instrumental
variables which are relevant to the question of instrument-exposure
specification. One such problem is exposure misspecification, for
example an instrument could associate with several closely related
putative outcomes, with only one of them actually having a causal effect
on the outcome. This problem has shown to be the case for SNPs
influencing different lipid fractions, for example {[}47,48{]}.

Mediation based network approaches, that go beyond analyses of two
variables, are very well established {[}37{]} and have a number of
extensions that make them valuable tools, including for example network
construction. But because they are predicated on the basic underlying
principles of mediation they are liable to suffer from the same issues
of measurement error. Recent advances in MR methodology, for example
applying MR to genetical genomics {[}49{]}, multivariate MR {[}48{]} and
mediation through MR {[}50--52{]} may offer more robust alternatives for
these more complicated problems.

\section{Methods}\label{methods}

\subsection{CIT test}\label{cit-test}

Here we describe how the CIT method {[}4{]} is implemented in the R
package \emph{R/cit} {[}18{]}. Assume an exposure \(x\) is instrumented
by a SNP \(g\), and the exposure \(x\) causes an outcome \(y\), as
described above. The following tests are then performed:

@@

The term in the 4th test can be rewritten as
\(cov(g, y|x) = cov(g, y - \hat{y})\) where
\(y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x\) is the residual
of \(y\) after adjusting for \(x\), and \(x\) is assumed to mediate the
association between the SNP and the outcome. The condition in the 4th
test is formulated as an equivalence testing problem that is estimated
using simulations, comparing the estimate from the data against
empirically obtained estimates for simulated variables where the
independence model is true (full details are given in {[}4{]}). We note
here that this approach is liable to fail, even when there is a true
causal relationship, when confounders of the exposure and outcome are
present, as these will induce collider bias.

@@

For the purposes of compiling simulation results we use an arbitrary
\(\alpha = 0.05\) value, though we stress that for real analyses it is
not good practice to rely on p-values for making causal inference, nor
is it reliable to depend on arbitrary significance thresholds {[}53{]}.

\subsection{MR causal test}\label{mr-causal-test}

Two stage least squares (2SLS) is a commonly used technique for
performing MR when the exposure, outcome and instrument data are all
available in the same sample. A p-value for this test, \(p_{MR}\), was
obtained using the \(systemfit\) function in the R package
\(R/systemfit\) {[}54{]}. Note that the value of \(p_{MR}\) is identical
when using the same genetic variant to instrument the influence of the
exposure \(x\) on the outcome \(y\), or erroneously, instrumenting the
outcome \(y\) on the exposure \(x\).

@@

To infer the direction of causality it is desirable to know which of the
variables, \(x\) or \(y\), is being directly influenced by the
instrument \(g\). This can be achieved by assessing which of the two
variables has the biggest absolute correlation with \(g\) (S2 Text),
formalised by testing for a difference in the correlations \(\rho_{gx}\)
and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
within a population {[}55{]}. It is calculated as

@@

An advantage of using the Steiger test in the two sample context is that
it can compare correlations in independent samples where sample sizes
are different. Steiger test statistics were calculated using the
\emph{r.test} function in the R package \emph{R/psych} {[}56{]}.

@@

\subsection{Causal direction sensitivity analysis for measurement
error}\label{causal-direction-sensitivity-analysis-for-measurement-error}

@@

\subsection{Simulations}\label{simulations}

@@

Two sample MR {[}30{]} was performed using summary statistics for
genetic influences on gene expression and DNA methylation. To do this we
obtained a list of 458 gene expression - DNA methylation associations as
reported in Shakhbazov et al {[}39{]}. These were filtered to be located
on the same chromosome, have robust correlations after correcting for
multiple testing, and to share a SNP that had a robust cis-acting effect
on both the DNA methylation probe and the gene expression probe. Because
only summary statistics were available (effect, standard error, effect
allele, sample size, p-values) for the instrumental SNP on the
methylation and gene expression levels, the Steiger test of two
independent correlations was used to infer the direction of causality
for each of the associations. The Wald ratio test was then used to
estimate the causal effect size for the estimated direction for each
association.

All analysis was performed using the R programming language {[}57{]} and
code is made available at
\url{https://github.com/explodecomputer/causal-directions} and
implemented in the MR-Base (\url{http://wwww.mrbase.org}) platform
{[}26{]}.

@@

\hypertarget{ref-Shakhbazov2016}{}
39. Shakhbazov K, Powell JE, Hemani G, Henders AK, Martin NG, Visscher
PM, et al. Shared genetic control of expression and methylation in
peripheral blood. BMC genomics. BioMed Central; 2016;17: 278.
doi:\href{https://doi.org/10.1186/s12864-016-2498-4}{10.1186/s12864-016-2498-4}

\hypertarget{ref-Bird2002}{}
40. Bird A. DNA methylation patterns and epigenetic memory. Genes \&
development. 2002;16: 6--21.
doi:\href{https://doi.org/10.1101/gad.947102}{10.1101/gad.947102}

\hypertarget{ref-Cole2014}{}
41. Cole DA, Preacher KJ. Manifest Variable Path Analysis: Potentially
Serious and Misleading Consequences Due to Uncorrected Measurement
Error. Psychological Methods. 2014;19: 300--315.
doi:\href{https://doi.org/10.1037/a0033805}{10.1037/a0033805}

\hypertarget{ref-Bose2014}{}
42. Bose M, Wu C, Pankow JS, Demerath EW, Bressler J, Fornage M, et al.
Evaluation of microarray-based DNA methylation measurement using
technical replicates: the Atherosclerosis Risk In Communities (ARIC)
Study. BMC Bioinformatics. 2014;15: 312.
doi:\href{https://doi.org/10.1186/1471-2105-15-312}{10.1186/1471-2105-15-312}

\hypertarget{ref-Bryant2011}{}
43. Bryant PA, Smyth GK, Robins-Browne R, Curtis N, Novak J, Sladek R,
et al. Technical Variability Is Greater than Biological Variability in a
Microarray Experiment but Both Are Outweighed by Changes Induced by
Stimulation. Khodursky AB, editor. PLoS ONE. Public Library of Science;
2011;6: e19556.
doi:\href{https://doi.org/10.1371/journal.pone.0019556}{10.1371/journal.pone.0019556}

\hypertarget{ref-Zhu2016}{}
44. Zhu Z, Zhang F, Hu H, Bakshi A, Robinson MR, Powell JE, et al.
Integration of summary data from GWAS and eQTL studies predicts complex
trait gene targets. Nature Genetics. Nature Research; 2016;48: 481--487.
doi:\href{https://doi.org/10.1038/ng.3538}{10.1038/ng.3538}

\hypertarget{ref-Mancuso2017}{}
45. Mancuso N, Shi H, Goddard P, Kichaev G, Gusev A, Pasaniuc B.
Integrating Gene Expression with Summary Association Statistics to
Identify Genes Associated with 30 Complex Traits. The American Journal
of Human Genetics. 2017;100: 473--487.
doi:\href{https://doi.org/10.1016/j.ajhg.2017.01.031}{10.1016/j.ajhg.2017.01.031}

\hypertarget{ref-Lee2013c}{}
46. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
association studies: quantification of power and genomic prediction
accuracy. PLoS One. 2013;8: e71494.
doi:\href{https://doi.org/10.1371/journal.pone.0071494\%20\%5Bdoi\%5D/rPONE-D-13-18776\%20\%5Bpii\%5D}{10.1371/journal.pone.0071494 {[}doi{]}\textbackslash{}rPONE-D-13-18776 {[}pii{]}}

\hypertarget{ref-Do2013}{}
47. Do R, Willer CJ, Schmidt EM, Sengupta S, Gao C, Peloso GM, et al.
Common variants associated with plasma triglycerides and risk for
coronary artery disease. Nature Genetics. Nature Research; 2013;45:
1345--1352. doi:\href{https://doi.org/10.1038/ng.2795}{10.1038/ng.2795}

\hypertarget{ref-Burgess2014a}{}
48. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
multivariable Mendelian randomization to disentangle the causal effects
of lipid fractions. PloS one. Public Library of Science; 2014;9:
e108891.
doi:\href{https://doi.org/10.1371/journal.pone.0108891}{10.1371/journal.pone.0108891}

\hypertarget{ref-Relton2012}{}
49. Relton CL, Davey Smith G. Two-step epigenetic Mendelian
randomization: a strategy for establishing the causal role of epigenetic
processes in pathways to disease. International journal of epidemiology.
2012;41: 161--76.
doi:\href{https://doi.org/10.1093/ije/dyr233}{10.1093/ije/dyr233}

\hypertarget{ref-Varbo2015}{}
50. Varbo A, Benn M, Smith GD, Timpson NJ, Tybjaerg-Hansen A,
Nordestgaard BG. Remnant cholesterol, low-density lipoprotein
cholesterol, and blood pressure as mediators from obesity to ischemic
heart disease. Circulation research. 2015;116: 665--73.
doi:\href{https://doi.org/10.1161/CIRCRESAHA.116.304846}{10.1161/CIRCRESAHA.116.304846}

\hypertarget{ref-Burgess2015}{}
51. Burgess S, Daniel RM, Butterworth AS, Thompson SG. Network Mendelian
randomization: using genetic variants as instrumental variables to
investigate mediation in causal pathways. International journal of
epidemiology. 2015;44: 484--95.
doi:\href{https://doi.org/10.1093/ije/dyu176}{10.1093/ije/dyu176}

\hypertarget{ref-Richmond2016}{}
52. Richmond RC, Hemani G, Tilling K, Davey Smith G, Relton CL.
Challenges and novel approaches for investigating molecular mediation.
Human molecular genetics. Oxford University Press; 2016;25: R149--R156.
doi:\href{https://doi.org/10.1093/hmg/ddw197}{10.1093/hmg/ddw197}

\hypertarget{ref-Sterne2001}{}
53. Sterne JAC, Smith GD. Sifting the evidence---what's wrong with
significance tests? BMJ. 2001;322: 226--231.

\hypertarget{ref-Henningsen2007}{}
54. Henningsen A, Hamann JD. systemfit : A Package for Estimating
Systems of Simultaneous Equations in R. Journal of Statistical Software.
2007;23: 1--40.
doi:\href{https://doi.org/10.18637/jss.v023.i04}{10.18637/jss.v023.i04}

\hypertarget{ref-Steiger1980}{}
55. Steiger JH. Tests for comparing elements of a correlation matrix.
Psychological Bulletin. 1980;87: 245--251.
doi:\href{https://doi.org/10.1037/0033-2909.87.2.245}{10.1037/0033-2909.87.2.245}

\hypertarget{ref-Revelle2015}{}
56. Revelle W. psych: Procedures for Psychological, Psychometric, and
Personality Research {[}Internet{]}. Evanston, Illinois: Northwestern
University; 2015. Available:
\url{http://cran.r-project.org/package=psych}

\hypertarget{ref-RCoreTeam2015}{}
57. R Core Team. R: A Language and Environment for Statistical Computing
{[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
2015. Available: \url{https://www.r-project.org/}

\section{Figures}\label{figures}
